---
title: "Reproducible Research: Peer Assessment 1"
author: "Sebastian Sanin-Restrepo"
date: "19/7/2019"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


## Loading and preprocessing the data
This section will load the datafile needed to study the activity monitoring. It also transform loaded file into data.table format, with date formatting for the *date* variable and numeric format for other variables.
```{r}
library(data.table)
dtable <- as.data.table(read.csv("activity.csv",header = T,colClasses = c("integer","Date","integer")))
head(dtable,n=5)
tail(dtable,n=5)

```
It can be seen that there are NA values for the *steps* variable both at the 5 first and last observations. We can also notice that *interval* format is an integer which consists on the hour and minutes of the measure. For instance, `r dtable$interval[58]` means that it is a measure of the steps at 4:45 A.M., also `r dtable$interval[1100]` is a measure at 19:35 P.M.

## What is mean total number of steps taken per day?

### Total step per day computation
In order to analyze the mean total number of steps per day we first must compute the total number of steps per day. In order to do that we use the *summarize* function joint with a *group_by* function.
```{r}
library(plyr)
library(dplyr)
dtable_noNAs <- na.omit(dtable)
steps_per_day <- group_by(dtable_noNAs,date) %>% summarise(total_steps=sum(steps,na.rm=T))
head(steps_per_day,n=10)
```
### Total step per day histogram
Looking for a more clear view of the total steps by day we show the following histogram of the previously computed variable (*steps_per_day$total_steps*). Histograms are useful because they provide us information about the frequency distribution of one variable that let us to look for extrange behaviors of a variable such as outlier values, heavy tail tendencies and bimodal distributions.

```{r}
with(steps_per_day,{
  hist(total_steps,breaks = 30, freq=F,col="lightblue",
       main="Total number of steps per day distribution",xlab="Number of steps in a day")
  lines(density(total_steps), col="red")
})
```

As it can be seen in the histogram, the empirical density distribution of the total number of steps per day seems similar to a normal distribution but with heavy tails. In terms of the bars, it can be seen that it has a very irregular distribution. It can be inferred that the median should be around 10000 and 11000 of steps per day.

### Mean and median computation
Finally, in order to really observe the meean and median computation we use the *summary* function.
```{r}
steps_per_day_stats <- with(steps_per_day,summary(total_steps))
steps_per_day_stats
```

The mean total steps per day is `r as.integer(round(steps_per_day_stats[4],0))` and the median `r as.integer(round(steps_per_day_stats[3],0))`. Both are almost the same suggesting that the distribution is very centered around those values.

## What is the average daily activity pattern?

### Time series pattern
The following time-series plot intention is to show patterns during the day in the 5-minutes time intervals. It is like an intra-daily seasonal plotting in order to look for intervals in which the number of steps is higher than another (for example, we could expect that during night the number of steps is much lower than between 6 a.m. and 9 a.m. that is when people are walking to work).

```{r}
library(ggplot2)
# Compute the interval-mean steps measure
steps_invervals <- aggregate(steps~interval,dtable_noNAs,mean)
# Plotting time series
ggplot(steps_invervals,aes(x=interval,y=steps))+geom_line()+
  labs(x="5-minute interval", y="Number of steps (mean of daily data)",
       title = "Number of steps by interval", 
       subtitle="Daily averages")+
  scale_x_continuous(breaks=seq(0, 2355, 200))+
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust=0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

From the plot it can be confirmed that the number of steps is much higher during the morning hours and it is around zero during night-time.

### The most intensive stepping interval
We can also find which is the interval that exhibits the maximum mean number of steps.
```{r}
max.interval <- with(steps_invervals,interval[which.max(steps)])
print(paste("The interval with the maximum mean number of steps is ",max.interval))
```

It turns out that the interval with more intensity is 8:35 a.m. confirming our expectations.

## Imputing missing values

### Missing values counting
The total missing values of the original data base could be computed in a straightforward way using the *summary* and *is.na* functions:
```{r}
summary(is.na(dtable$steps))
```

We can see that there are 2304 missing values.

### Filling missing values and new dataset
In order to fill missing values the strategy followed consists on using daily mean value per interval. It follows the following steps:
  1 - Compute the mean value through days by interval on non-missing observations.
  2 - Replace NA value of a given observation by its interval mean value.

For the first step we use the same code used before for compute interval mean through days:
```{r}
# Compute mean by interval column
steps_invervals <- aggregate(steps~interval,dtable_noNAs,mean)
# Join tables
ddtable <- as.data.table(join(dtable,steps_invervals,by="interval"))
colnames(ddtable) <- c("steps","date","interval","steps_fill")

# Create new database with NAs filled
steps_filled <- ddtable
steps_filled$steps <- with(steps_filled,ifelse(is.na(steps),steps_fill,steps))
steps_filled <- steps_filled[,-("steps_fill")]
```
Finally the *steps_filled* is a new database that has three variables:

 - date: date in YYYY-MM-DD format.
 - interval: 5-minute interval value.
 - steps: total number of steps with NAs filled with the average for the specific interval.

### Histograms and computations
Now we are going to create an histogram and compute some descriptive statistics using the new database with no NAs values.
```{r}
# Create new summary datatable
steps_fill_per_day <- group_by(steps_filled,date) %>% summarise(total_steps=sum(steps,na.rm=T))

# Create histogram
with(steps_fill_per_day,{
  hist(total_steps,breaks = 30, freq=F,col="lightblue",
       main="Total number of steps per day distribution",xlab="Number of steps in a day")
  lines(density(total_steps), col="red")
})

# Compute summary statistics
steps_fill_per_day_stats <- with(steps_fill_per_day,summary(total_steps))
steps_fill_per_day_stats
```
The mean total steps per day is `r as.integer(round(steps_fill_per_day_stats[4],0))` and the median `r as.integer(round(steps_fill_per_day_stats[3],0))`. Before computed values were `r as.integer(round(steps_per_day_stats[4],0))` and the median `r as.integer(round(steps_per_day_stats[3],0))` respectively. Both are almost the same, possibly due to the fact that mean values per interval were applied to each missing value making it to converve to the mean and median value.


## Are there differences in activity patterns between weekdays and weekends?
### Creating the factor variable
This part will be in charge of creating the factor variable called weekend, that will assign 1 to weekends and 0 in other case.
```{r}
# Change local time (because my country is Spanish based)
Sys.setlocale("LC_TIME", "English")

steps_filled[,weekend:=ifelse(weekdays(date)=="Sunday"|weekdays(date)=="Saturday",1,0)]

# Converting to factor variables with descriptive labels
steps_filled$weekend <-factor(steps_filled$weekend,labels = c("Weekday","Weekend"))
```
### Plotting time
```{r}
# Compute the weekend/weekday-mean steps measure
steps_filled_av <- aggregate(steps~interval+weekend,steps_filled,mean)
library(lattice)
xyplot(steps~interval|weekend,data=steps_filled_av,layout=c(1,2),
       type="l",xlab = "5-minute interval",ylab="Average number of steps",
       main="Step Pattern difference between Weekdays and Weekend")
```

From the previous plot it can be inferred that the weekends exhibits less steps than the weekdays. On the other hand, weekend steps are less dispersed than for the weekday case.